PAMELA: an annotation-based Java Modeling Framework

### 2021-03-23 6-Experimentations.tex
review up to p5 (details only)

The full Openflexo infrastructure is composed of about 50 components. In most components, PAMELA framework is extensively used. As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes. 159 of those classes (mostly the diagraming model) are defined as Pamela \texttt{ModelEntity}.

The total base of code for Openflexo infrastructure represents around
\num{900000} lines of Java code. Regarding backend modeling, the PAMELA
framework is extensively used in Openflexo core as well as in most technology
adapters, with very few specific implementations for properties. The
observer/observable pattern is generally used for graphical user interfaces,
which also rely on PAMELA framework.

An interesting experimentation has been done in the context of Openflexo development process. When Pamela has been integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to Pamela. Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration. In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.

Pamela implementation is now really stable and mature. According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of models and code is greatly improved compared to a "code generation"-based solution.

@@

The main results of the project are a requirements modeling multi-view
language, its associated design process and the development of an open-source
platform called Formod\cite{FormodWebSite}, built using Openflexo
infrastructure and PAMELA framework. %\todo{Say how and where PAMELA was used}
The requirements modeling language is based on KAOS~\cite{kaos} for goal
modeling and SysML for the structural part of a system. The associated domain
modeling language, used to describe system domain knowledge, extends the two
ontology languages OWL~\cite{owl} and PLIB~\cite{plib}. The graphical notations
are then translated into Event-B~\cite{eventb}, a formal specification method supported by verification tools.

>>>>

The full Openflexo infrastructure is composed of about 50 components. In most components, the PAMELA framework is extensively used. As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes. 159 of those classes (mostly the diagraming model) are defined as Pamela \texttt{ModelEntity}.

The total base of code for Openflexo infrastructure represents around
\num{900000} lines of Java code. Regarding backend modeling, the PAMELA
framework is extensively used in Openflexo core as well as in most technology
adapters, with very few specific implementations for properties. The
observer/observable pattern is generally used for graphical user interfaces,
which also rely on the PAMELA framework.

An interesting experimentation has been done in the context of Openflexo development process. When PAMELA has been integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to PAMELA. Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration. In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.

The PAMELA implementation is now really stable and mature. According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no PAMELA-specific issue, while co-evolution of models and code is greatly improved compared to a "code generation"-based solution.

@@

The main results of the project are a requirements modeling multi-view
language, its associated design process and the development of an open-source
platform called Formod\cite{FormodWebSite}, built using Openflexo
infrastructure and the PAMELA framework. %\todo{Say how and where PAMELA was used}
The requirements modeling language is based on KAOS~\cite{kaos} for goal
modeling and SysML for the structural part of a system. The associated domain
modeling language, used to describe system domain knowledge, extends the two
ontology languages OWL~\cite{owl} and PLIB~\cite{plib}. The graphical notations
are then translated into Event-B~\cite{eventb}, a formal specification method supported by verification tools.

### 2021-03-23 Todo-review2.tex
review up to p5 (details only)

There remain or are introduced however, a substantial number of typos, language errors, and other minor issues that need correcting:
- "to /avoid" on p3 (DONE)
- "detailed thereafter" -> "detailed hereafter" (It's Correct)
- "increasing PAMELA metamodel" -> "by _extending_ PAMELA_'s_ metamodel" (DONE)
- "with setter/getter": make both plural (DONE)
- POJO used here and elsewhere yet the acronym is introduced much later (DONE)
- p4 l11-12: reduce etc. should be reduce_s_ etc. (Keep as it is)
- odd spacing before : consistently throughout bullet list + text  (Keep as it is: I trust Latex)
- "in the section 3.2" -> "in Section 3.2" (style: named references deserve a capital and no article before) (DONE)
- p5 and many times: "PAMELA metamodel", "PAMELA model", "PAMELA framework" need "the" before them.
- "defines API": insert "an"
- "in Java interface": insert "a"
- "returning desired value", "taking value" etc.: insert "the"
- "May also declares"...
- p6 l46: @Embedded/"as embedded" meaning?
- p7 "of launched Java application" missing "the"
- "is/are" -> "is (are)", in line with parentheses earlier in sentence
- "Experimented programmers": I doubt it, unless you got major ethical clearance
- "Listing 2: model" -> "~: Model" consistent with Listing 1
- the book title is "Lord of the ring_s_" with s
- "using classical Java ... scheme": missing "the"
- "Java programming language": insert "The"
- l43 inconsistent spacing
- l46 package ... package
- p11 l36 sticking into margin
- S.4.3 please add size in LoC and classes
- java -> Java
- p13 "diana component": insert "the"
- "experimentation" -> "experiment"
- "When Pamela has been" -> "After Pamela was"
- l36 the quotation mark before code generation is the wrong one
- p 14 "appertaining" -> "pertaining", "relating"

@@

Minors:
- POJO should be explain the first time p3 not the second time.
- p5 it is written twice that ModelEntities are encoded in a Java Interface
- p13, it is written twice that Pamela framework is extensively used.

>>>>

There remain or are introduced however, a substantial number of typos, language errors, and other minor issues that need correcting:
- "to /avoid" on p3 (DONE)
- "detailed thereafter" -> "detailed hereafter" (It's Correct)
- "increasing PAMELA metamodel" -> "by _extending_ PAMELA_'s_ metamodel" (DONE)
- "with setter/getter": make both plural (DONE)
- POJO used here and elsewhere yet the acronym is introduced much later (DONE)
- p4 l11-12: reduce etc. should be reduce_s_ etc. (Keep as it is)
- odd spacing before : consistently throughout bullet list + text  (Keep as it is: I trust Latex)
- "in the section 3.2" -> "in Section 3.2" (style: named references deserve a capital and no article before) (DONE)
- p5 and many times: "PAMELA metamodel", "PAMELA model", "PAMELA framework" need "the" before them. (Done ; with Pamela -> PAMELA ; to be checked)
- "defines API": insert "an" (DONE)
- "in Java interface": insert "a" (DONE)
- "returning desired value", "taking value" etc.: insert "the" (DONE)
- "May also declares"... (DONE)
- p6 l46: @Embedded/"as embedded" meaning?
- p7 "of launched Java application" missing "the"
- "is/are" -> "is (are)", in line with parentheses earlier in sentence
- "Experimented programmers": I doubt it, unless you got major ethical clearance
- "Listing 2: model" -> "~: Model" consistent with Listing 1
- the book title is "Lord of the ring_s_" with s
- "using classical Java ... scheme": missing "the"
- "Java programming language": insert "The"
- l43 inconsistent spacing
- l46 package ... package
- p11 l36 sticking into margin
- S.4.3 please add size in LoC and classes
- java -> Java
- p13 "diana component": insert "the"
- "experimentation" -> "experiment"
- "When Pamela has been" -> "After Pamela was"
- l36 the quotation mark before code generation is the wrong one
- p 14 "appertaining" -> "pertaining", "relating"

@@

Minors:
- POJO should be explain the first time p3 not the second time. (DONE)
- p5 it is written twice that ModelEntities are encoded in a Java Interface (DONE)
- p13, it is written twice that Pamela framework is extensively used.

### 2021-03-23 architecture.tex
review up to p5 (details only)

A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}. A \texttt{ModelEntity} reflects a concept and is encoded in a Java \texttt{interface}. PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities. Note that reification of \texttt{ModelEntity} is performed in a Java \texttt{interface} (and not a class), which only defines API without any implementation for methods. 
Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which
will be responsible for providing custom code implementing domain or
application-specific behavior. A partially implemented \texttt{abstract} Java
\texttt{class} may be defined as partial base implementation (conforms to
implemented interface), as default methods in Java interfaces would be
(since Java 8).

A \texttt{ModelEntity} also defines some properties, encoded as \texttt{ModelProperty}. A \texttt{ModelProperty} is identified by a name, a cardinality (simple or multiple) and a type, which can be a reference to another \texttt{ModelEntity}, or a Java type (a primitive or an arbitrary complex Java type). Depending on its cardinality, a \texttt{ModelProperty} is bound to a set of methods reflecting use of property.
\begin{itemize}
    \item A \emph{read-only single property} will define read access of its value using a \emph{getter} (a Java method defined in Java interface taking no argument and returning desired value).
    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a Java method taking value to be set as unique argument).
    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a Java method taking value to be added as unique argument), a \emph{remover} (a Java method taking value to be removed as unique argument), and may define additional methods for extended features such as re-indexing, for example.
\end{itemize}

@@

\begin{itemize}
    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May declare an abstract entity.
    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract Java \texttt{class} to be used as base implementation.
    \item \texttt{@Implementation}: tag annotating a partial implementation (abstract inner \texttt{class} defined in implemented \texttt{interface}), and used in the context of multiple inheritance.
    \item \texttt{@Getter(String)}: tag annotating method as unique getter for implicit \emph{ModelProperty} whose identifier is the declared String value. May also declares cardinality, eventual inverse property, default value and some other features.
    \item \texttt{@Setter(String)}: tag annotating method as unique setter for implicit \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Adder(String)}: tag annotating method as unique adder for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Remover(String)}: tag annotating method as unique remover for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Reindexer(String)}: tag annotating method as unique re-indexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}.
   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}.
    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute.
    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}.
    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as
      embedded. %according to PAMELA semantics.
    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model.
    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances.

>>>>

A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}. A \texttt{Model\-Entity} reflects a concept and is encoded in a Java \texttt{interface}. 
Then, a \texttt{Model\-Entity} only defines an API (Application Programming Interface) without any implementation for methods. 
Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which
will be responsible for providing custom code implementing domain or
application-specific behavior. A partially implemented \texttt{abstract} Java
\texttt{class} may be defined as partial base implementation (conforms to
implemented interface), as default methods in Java interfaces would be
(since Java 8).
The PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities.

A \texttt{ModelEntity} also defines some properties, encoded as \texttt{ModelProperty}. A \texttt{ModelProperty} is identified by a name, a cardinality (simple or multiple) and a type, which can be a reference to another \texttt{ModelEntity}, or a Java type (a primitive or an arbitrary complex Java type). Depending on its cardinality, a \texttt{ModelProperty} is bound to a set of methods reflecting use of property.
\begin{itemize}
    \item A \emph{read-only single property} will define read access of its value using a \emph{getter} (a Java method defined in Java interface taking no argument and returning the desired value).
    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a Java method taking the value to be set as unique argument).
    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a Java method taking the value to be added as unique argument), a \emph{remover} (a Java method taking the value to be removed as unique argument), and may define additional methods for extended features such as re-indexing, for example.
\end{itemize}

@@

\begin{itemize}
    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May also declare an abstract entity.
    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract Java \texttt{class} to be used as base implementation.
    \item \texttt{@Implementation}: tag annotating a partial implementation (abstract inner \texttt{class} defined in implemented \texttt{interface}), and used in the context of multiple inheritance.
    \item \texttt{@Getter(String)}: tag annotating method as unique getter for implicit \emph{ModelProperty} whose identifier is the declared String value. May also declares cardinality, eventual inverse property, default value and some other features.
    \item \texttt{@Setter(String)}: tag annotating method as unique setter for implicit \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Adder(String)}: tag annotating method as unique adder for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Remover(String)}: tag annotating method as unique remover for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Reindexer(String)}: tag annotating method as unique re-indexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}.
   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}.
    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute.
    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}.
    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as
      embedded. %according to PAMELA semantics.
    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model.
    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances.

### 2021-03-23 implementation.tex
review up to p5 (details only)

From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the Java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to the underlying model to be executed. This also provides an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at runtime (support for Design by Contract, aka JML).

@@

PAMELA implementation is modularized.

@@

A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developed to help measure that overhead. We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value). From this simple model, two java implementations have been derived using code generation. The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpreted (it uses PAMELA framework and defines only interfaces and API methods).

@@

Moreover, comparison between the two implementations should be tempered by the fact that PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.).

>>>>

From a technical point of view, the PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the Java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to the underlying model to be executed. This also provides an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at runtime (support for Design by Contract, aka JML).

@@

The PAMELA implementation is modularized.

@@

A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developed to help measure that overhead. We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value). From this simple model, two java implementations have been derived using code generation. The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpreted (it uses the PAMELA framework and defines only interfaces and API methods).

@@

Moreover, comparison between the two implementations should be tempered by the fact that the PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.).
